---
title: "Class 10: Halloween Mini Project"
author: "Jiayi Zhou (PID:A17856751)"
format: pdf
toc: true
---

As it is nearly Halloween and the half way point in the quarter let's do a mini project to help us figure out the best candy!

Our come from the 538 website and is available as a CSV file:

# Data Import

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```

```{r}
flextable::flextable(head(candy, 10))
```

>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

```{r}
library(tidyverse)
```

```{r}
candy %>%
  nrow()
```
> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite winpercent
```{r}
candy["Hershey's Milk Chocolate", ]$winpercent
```

```{r}
library(dplyr)

candy |>
  filter(rownames(candy) == "Hershey's Milk Chocolate") |>
  select(winpercent)
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy |>
  filter(rownames(candy) == "Kit Kat") |>
  select(winpercent)
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy |>
  filter(rownames(candy) == "Tootsie Roll Snack Bars") |>
  select(winpercent)
```

## Quick Overview of the dataset

```{r}
library("skimr")
skimr::skim(candy)
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent is on 0-100 scale and the rest are 0-1 scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

That the candy does not contain chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=20)
```
```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_density()
```

> Q9. Is the distribution of winpercent values symmetrical?

Based on the summary statistics, the mean (50.32) is greater than the median (47.83), and the maximum (84.18) is much farther from the median than the minimum (22.45). 
The distribution is not symmetrical; it is right-skewed (positively skewed)

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```
```{r}
summary(candy$winpercent)
```

The median winpercent is 47.83, which is below 50%. Therefore, the center of the distribution is below 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Chocolate is higher ranked on average.

```{r}
# 1. Find all chocolate candy in the dataset.
# 2, Find their winpercent values
# Calculate the mean of these values

# 4-6. Do the same for fruity candy
# 7. Compare mean winpercents of chocolate vs fruity
# 8. Pick the highest as the winner

choc.inds <- candy$chocolate==1
choc.win <- candy[choc.inds,]$winpercent
choc.win <- mean(choc.win)
choc.win
```

```{r}
mean(candy[candy$chocolate==1,]$winpercent)
```

```{r}
mean(candy[candy$fruity==1,]$winpercent)

fruit.ind <- candy$fruity==1
fruit.win <- candy[fruit.ind,]$winpercent
fruit.mean <- mean(fruit.win)

choc.ind <- candy$chocolate==1
choc.win <- candy[choc.ind,]$winpercent
choc.mean <- mean(choc.win)
```
```{r}
candy |> 
  filter(chocolate == 1) |> 
  select(winpercent)
```

> Q12. Is this difference statistically significant?

Yes, the difference is statistically significant.

```{r}
t.test(choc.win, fruit.win)
```
> Q13. What are the five least liked candy types in this set?

```{r}
candy |>
  arrange(winpercent) |>
  head(5)
```
```{r}
x <- c(5,1,10,4)
sort(x)
#sort(x)
#(candy$wonpercent)
```
```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,],5)
```
> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(-winpercent) |>
  head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) + 
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent)) + 
  geom_col()
```
Add some colors based on the "type of candy"

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$fruity)] = "pink"
my_cols[as.logical(candy$bar)] = "brown"
my_cols
```

```{r}
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent)) + 
  geom_col(fill=my_cols)
```

Now, for the first time, using this plot we can answer questions like:

> Q17. What is the worst ranked chocolate candy?

```{r}
candy |> 
  dplyr::filter(chocolate == 1) |> 
  dplyr::slice_min(winpercent, n = 1, with_ties = FALSE)
```

Sixlets.

> Q18. What is the best ranked fruity candy?

```{r}
candy |> 
  dplyr::filter(fruity == 1) |> 
  dplyr::slice_max(winpercent, n = 1, with_ties = FALSE)
```
Starburst.

## Winpercent and Pricepercent

A plot with both variables/columns winpercent and pricepercent

```{r}
my_cols[as.logical(candy$fruity)] = "red"

ggplot(candy) +
  aes(x=winpercent, y=pricepercent, label=row.names(candy)) +
  geom_point(col=my_cols) +
  geom_text()
```

```{r}
library(ggrepel)

ggplot(candy) +
  aes(x=winpercent, y=pricepercent, label=row.names(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(max.overlaps = 7)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
candy |>
  dplyr::mutate(type = dplyr::case_when(chocolate == 1 & bar == 1 ~ "chocolate bar",
                                        chocolate == 1 & bar == 0 ~ "chocolate other",
                                        fruity == 1 ~ "fruity",
                                        TRUE ~ "other")) |>
  dplyr::group_by(type) |>
  dplyr::summarise(mean_win = mean(winpercent, na.rm = TRUE),
                   mean_price = mean(pricepercent, na.rm = TRUE), .groups = "drop") |>
  dplyr::mutate(score = dplyr::min_rank(mean_price) + dplyr::min_rank(dplyr::desc(mean_win))) |>
  dplyr::slice_min(score, n = 1, with_ties = FALSE)
```
Fruity type is the highest ranked in terms of winpercent for the least money.

```{r}
candy |> 
  dplyr::mutate(score = dplyr::min_rank(pricepercent) +
                  dplyr::min_rank(dplyr::desc(winpercent))) |> 
  dplyr::slice_min(score, n = 1, with_ties = FALSE)
```
Reese's Miniatures is the highest single candy ranked in terms of winpercent for the least money

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
candy |>
  arrange(-pricepercent) |>
  head(5)
```
```{r}
candy |>
  arrange(-pricepercent) |>
  head(5)|> 
  dplyr::slice_min(winpercent, n = 1, with_ties = FALSE)
```
Nik L Nip, Ring Pop, Nestlé Smarties, Hershey's Krackel, Hershey’s Milk Chocolate.
Among these, the least popular (lowest winpercent) is Nik L Nip.

## Exploring the correlation structure

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
cij
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolaty and Fruity, bar and pluribus. etc.

> Q23. Similarly, what two variables are most positively correlated?

Chocolaty and bar, fruity and hard, etc.

## Principle Componenet Analysis

The function to use is called `prcomp()` with an optional `scale=T/F` argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
Our main PCA result figure

```{r}
ggplot(pca$x, aes(PC1, PC2)) + 
  geom_point(col=my_cols) + 
  geom_text_repel(aes(label = rownames(pca$x)), col = my_cols, max.overlaps = 5)
```

Interactive plots that can be zoomed on and "brushed" over can be made with the **plotly** package, It's output is interactive and will not render to pdf :-(

```{r}
#install.packages("plotly")
library(plotly)
#plotly(p)
```

We should also examine the variable "loadings" or contributions of the original variables to the new PCs

```{r}
ggplot(pca$rotation)+
  aes(PC1, row.names(pca$rotation))+
  geom_col()
```
> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

fruity, pluribus, and hard load strongly positive on PC1. 

PC1 is separating non-chocolate/fruit-style, hard, multipack candies on the positive side from chocolate/bar/price/winpercent features on the negative side—matching the known opposition between chocolate and fruity and the similarity between chocolate and bar.
